---
title: "Using msigdf with enrichment tools"
author:
- name: Enrique M. Toledo
  email: enriquetoledo@gmail.com
package: msigdf
output:
  html_document:
    theme:
      version: 5
      bootswatch: flatly
    highlight: monochrome
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    css: styles.css
abstract: |
  Recipes for reshaping the `msigdf` tibbles into the formats expected by common
  gene set enrichment tools: `fgsea`, `clusterProfiler`, and anything else that
  wants either a named list of gene sets or a two-column `TERM2GENE` frame.
vignette: |
  %\VignetteIndexEntry{Using msigdf with enrichment tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE)
```

```{r}
library(dplyr)
library(msigdf)
```

# Scope

`msigdf` ships the gene sets only. It deliberately does not depend on any
enrichment package, so `fgsea` and `clusterProfiler` are **not** installed
alongside it.

The chunks that reshape `msigdf` data are evaluated below, so the output you see
is real. The chunks that call an enrichment package are marked `eval=FALSE` and
are shown as recipes to copy — install the relevant Bioconductor package first:

```{r, eval=FALSE}
# install.packages("BiocManager")
BiocManager::install(c("fgsea", "clusterProfiler"))
```

# A named list of gene sets

`fgsea::fgsea()` and several other tools expect gene sets as a named list: one
element per gene set, each holding a character vector of gene identifiers.
`split()` produces exactly that.

Using the human hallmark collection (`category_code == "h"`):

```{r}
hallmark <- msigdf.human %>%
  filter(category_code == "h")

pathways <- split(hallmark$symbol, hallmark$geneset)

length(pathways)
lengths(pathways) %>% head()
pathways[["HALLMARK_NOTCH_SIGNALING"]] %>% head(10)
```

Feed that straight to `fgsea`, together with your own named vector of ranking
statistics:

```{r, eval=FALSE}
library(fgsea)

# `stats` is a named numeric vector: names are gene symbols, values are the
# ranking metric (log fold change, signed -log10 p-value, ...).
res <- fgsea(pathways = pathways, stats = stats)

res %>%
  as_tibble() %>%
  arrange(padj) %>%
  select(pathway, pval, padj, NES, size)
```

Any collection works the same way. To restrict to one sub-collection, filter on
`category_subcode` as well:

```{r}
reactome <- msigdf.human %>%
  filter(category_code == "c2", category_subcode == "cp.reactome")

reactome_pathways <- split(reactome$symbol, reactome$geneset)
length(reactome_pathways)
```

# A TERM2GENE data frame

`clusterProfiler::enricher()` and `GSEA()` take a two-column data frame: gene set
name first, gene identifier second. That is the `msigdf` layout minus the
collection columns.

```{r}
term2gene <- msigdf.human %>%
  filter(category_code == "h") %>%
  select(term = geneset, gene = symbol)

term2gene
```

Then:

```{r, eval=FALSE}
library(clusterProfiler)

# Over-representation on a character vector of gene symbols.
ora <- enricher(gene = my_genes, TERM2GENE = term2gene)

# Or GSEA on a sorted, named numeric vector.
gsea <- GSEA(geneList = sorted_stats, TERM2GENE = term2gene)
```

# Mouse

The mouse tibble has the same shape. Note the collection codes differ: hallmark
is `mh` rather than `h`, and the numbered collections run `m1`-`m8`.

```{r}
msigdf.mouse %>%
  distinct(category_code) %>%
  arrange(category_code)
```

```{r}
mouse_hallmark <- msigdf.mouse %>%
  filter(category_code == "mh")

mouse_pathways <- split(mouse_hallmark$symbol, mouse_hallmark$geneset)

length(mouse_pathways)
mouse_pathways[["HALLMARK_NOTCH_SIGNALING"]] %>% head(10)
```

Mouse gene sets carry mouse symbols, so no ortholog mapping is needed if your
data is already mouse.

# Linking results back to MSigDB

`msigdf.urls` and `msigdf.mouse.urls` map every gene set name to its page on the
MSigDB website. Join enrichment output against them to get clickable links in a
report.

```{r}
top_sets <- msigdf.human %>%
  filter(category_code == "h") %>%
  count(geneset, name = "n_genes") %>%
  arrange(n_genes) %>%
  head(5)

top_sets %>%
  inner_join(msigdf.urls, by = "geneset") %>%
  select(geneset, n_genes, url)
```

The same join applied to an `fgsea` result:

```{r, eval=FALSE}
res %>%
  as_tibble() %>%
  arrange(padj) %>%
  head(20) %>%
  inner_join(msigdf.urls, by = c("pathway" = "geneset")) %>%
  select(pathway, padj, NES, url)
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
