---
title: "MSigDF: Molecular Signature Database (MSigDB) in a Data Frame"
author:
- name: Stephen D. Turner 
  email: vustephen@gmail.com
- name: Enrique M. Toledo 
  email: enriquetoledo@gmail.com
maintenace: Enrique M. Toledo (enriquetoledo@gmail.com)
package: msigdf
output:
  html_document:
    theme:
      version: 5
      bootswatch: flatly
    highlight: monochrome
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    css: styles.css
abstract: |
  This data package contains the Molecular Signature Database (MSigDB) for both human and mouse in separate data frames (tibbles).
  Each data frame (`msigdf.human` and `msigdf.mouse`) contains four columns: the collection code (`h` or `c1`-`c9` for human, `mh` or `m1`-`m8` for mouse),
  the sub-collection code, the gene set name, and the gene symbols in that set. The `msigdf.urls` and `msigdf.mouse.urls` tibbles contain links to
  descriptions of each gene set on the MSigDB website.
  **[Source code available on GitHub](https://github.com/ToledoEM/msigdf)**.
vignette: |
  %\VignetteIndexEntry{MSigDF: Molecular Signature Database (MSigDB) in a Data Frame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE)
```

# Data sources

Original data from the Broad Institute's Molecular Signature Database (MSigDB)^[<http://www.broad.mit.edu/gsea/msigdb/index.jsp>], redistributed as separate gmt data files from the MSigDB.

----


**_Procedure:_** 

_1. The current MSigDB v2026.1 gmt files were downloaded from Broad ftp._      
_2. This was done with the human and mouse gene sets_     
_3. Each collection was converted to a list in R, and written to a RData file using `save()`._        

----

See the script in `data-raw/` to see how the data frames (tibbles) were created.

# Example usage

There are four data frames (tibbles) in this package. The `msigdf.human` data frame has columns for each MSigDB collection divided by sub-collection (like cc, bp and mf for C5). The format of the data is tidy, so each row is a single gene set collection, sub-collection and gene symbol. The `msigdf.mouse` data frame has the same structure for the mouse collections. The `msigdf.urls` and `msigdf.mouse.urls` data frames link the name of each gene set to its page on the MSigDB website.

New C5 ontology information was added to the category subcode for easy filtering and consistency.   

- HPO: Human Phenotype Ontology
- MF: GO Molecular Function ontology
- BP: GO Biological Process ontology
- CC: GO Cellular Component ontology


The data sets in this package have several million rows. The package imports the tibble package so they're displayed nicely.

```{r}
library(dplyr)
library(tibble)
library(purrr)
library(msigdf)
```

Take a look:

```{r}
msigdf.human %>% head()
msigdf.mouse %>% head()
msigdf.urls %>% as.data.frame() %>% head()
```

Just get the entries for the [KEGG non-homologous end joining pathway](http://www.genome.jp/kegg/pathway/hsa/hsa03450.html): 

```{r}
msigdf.human %>% 
  filter(geneset=="KEGG_NON_HOMOLOGOUS_END_JOINING")
```

Some software, e.g., [fGSEA](https://github.com/ctlab/fgsea) might require gene sets to be a named list of genes identifiers, where the name of each element in the list is the name of the pathway. This is how the data was originally structured, and we can return to it with `plyr::dlply()`. Here, let's use only the hallmark sets, and after we `dlply` the data into this named list format, get just the first few pathways, and in each of those, just display the first few gene symbols.

```{r}
msigdf.human %>%
  filter(category_code=="h") %>%
  select(geneset, symbol) %>%
  group_by(geneset) %>%
  summarize(symbol=list(symbol)) %>%
  deframe() %>%
  head() %>%
  map(head)
```

# Further exploration

The number of gene sets in each collection for each organism is dependent of the construction at MSigDB.

**Human Collection of gene sets**
https://www.gsea-msigdb.org/gsea/msigdb/human/collections.jsp

```{r}
msigdf.human %>%
  group_by(category_code,category_subcode) %>% 
  tally()
```

**Mouse Collection of gene sets**
https://www.gsea-msigdb.org/gsea/msigdb/mouse/collections.jsp

```{r}
msigdf.mouse %>%
  group_by(category_code,category_subcode) %>% 
  tally()
```

Get the URL for the hallmark set with the fewest number of genes (Notch signaling). Optionally, `%>%` this to `browseURL` to open it up in your browser.

```{r}
msigdf.human %>%
  filter(category_code=="h") %>%
  count(geneset) %>%
  arrange(n) %>%
  head(1) %>%
  inner_join(msigdf.urls, by="geneset") %>%
  pull(url)
```

Just look at the number of genes in each KEGG pathway (sorted descending by the number of genes in that pathway):

```{r}
msigdf.human %>%
  filter(category_code=="c2" & grepl("^KEGG_", geneset)) %>%
  count(geneset) %>% 
  arrange(desc(n))
```


# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
